Channel Shaping Using Reconfigurable Intelligent Surfaces: From Diagonal Model to Beyond

### 2024-05-01 main.tex
:sparkles: Add complexity analysis

	% 	\item What is the entire achievable singular value region of a \gls{mimo} channel with \gls{ris}?

@@

	% pioneer
	Third, we quantify the capability of a \gls{bd}-\gls{ris} to redistribute the singular values of a point-to-point \gls{mimo} channel.
	The {Pareto frontiers} are characterized by optimizing the {weighted sum of singular values}, where the weights can be positive, zero, or negative.
	This problem is solved by the proposed geodesic \gls{rcg} algorithm.
	The resulting achievable singular value region generalizes most relevant metrics and provides an intuitive channel shaping benchmark.
	% We also and derive some analytical singular value bounds for rank-deficient \gls{mimo} and unitary \gls{ris}.
	To explore the shaping limits of different \gls{ris} architectures, we also derive individual and collective singular value bounds for rank-deficient channels and unitary \gls{ris} via matrix analysis.
	Results validate those bounds and show that increasing \gls{bd}-\gls{ris} group size significantly enlarges the singular value region, providing wider dynamic range and better trade-off.
	% dynamic range and trade-off of singular values are significantly improved by \gls{bd}-\gls{ris}, especially in large-scale \gls{mimo} systems.
	This is the first work to comprehensively answer the channel shaping capability question from both numerical and analytical perspectives in the most general \gls{bd}-\gls{ris} setup.
	% from a Pareto perspective.

@@

		A Lie group is simultaneously a continuous group and a differentiable manifold.
		Lie algebra refers to the tangent space of the Lie group at the identity element.
		The exponential map acts as a bridge between the Lie algebra and Lie group, which allows one to recapture the local group structure using linear algebra techniques.
		The set of unitary matrices $\mathbb{U}^{L \times L}$ forms a Lie group $U(L)$ under multiplication, and the corresponding Lie algebra $\mathfrak{u}(L) \triangleq \mathcal{T}_{\mathbf{I}}\mathbb{U}^{L \times L} = \{\mathbf{M} \in \mathbb{C}^{L \times L} \mid \mathbf{M}^\mathsf{H} + \mathbf{M} = \mathbf{0}\}$ consists of skew-Hermitian matrices.
		A geodesic emanating from the identity with velocity $\mathbf{D} \in \mathfrak{u}(L)$ can be described by \cite{Edelman1998}
		\begin{equation}
			\mathbf{G}_\mathbf{I}(\mu) = \exp(\mu \mathbf{D}),
			\label{eq:geodesic_identity}
		\end{equation}
		where $\exp(\mathbf{A}) = \sum_{k=0}^\infty (\mathbf{A}^k/k!)$ is the matrix exponential and $\mu$ is the step size (i.e., magnitude of the tangent vector).
		Note that the right translation is an isometry in $U(L)$.
		During the optimization of group $g$, the geodesic evaluated at the identity \eqref{eq:geodesic_identity} should be translated to $\mathbf{\Theta}_g^{(r)}$ for successive updates \cite{Abrudan2008}
		\begin{equation}
			\mathbf{G}_g^{(r)}(\mu) = \mathbf{G}_\mathbf{I}(\mu) \mathbf{\Theta}_g^{(r)} = \exp(\mu \mathbf{D}_g^{(r)}) \mathbf{\Theta}_g^{(r)},
			\label{eq:geodesic_translated}
		\end{equation}
		while the Riemannian gradient evaluated at $\mathbf{\Theta}_g^{(r)}$ \eqref{eq:gradient_riemannian} should be translated back to the identity for exploiting the Lie algebra \cite{Abrudan2008}
		\begin{equation}
			\tilde{\nabla}_{\mathrm{R},g}^{(r)} = \nabla_{\mathrm{R},g}^{(r)} \mathbf{\Theta}_g^{(r)\mathsf{H}} = \nabla_{\mathrm{E},g}^{(r)} \mathbf{\Theta}_g^{(r)\mathsf{H}} - \mathbf{\Theta}_g^{(r)} {\nabla_{\mathrm{E},g}^{(r)\mathsf{H}}}.
			\label{eq:gradient_translated}
		\end{equation}
		After gradient translation, the deviation parameter and conjugate direction can be determined similarly to \eqref{eq:parameter_cg} and \eqref{eq:direction_cg}
		\begin{equation}
			\tilde{\gamma}_g^{(r)} = \frac{\mathrm{tr}\bigl((\tilde{\nabla}_{\mathrm{R},g}^{(r)} - \tilde{\nabla}_{\mathrm{R},g}^{(r-1)}) {\tilde{\nabla}_{\mathrm{R},g}^{(r)\mathsf{H}}}\bigr)}{\mathrm{tr}\bigl(\tilde{\nabla}_{\mathrm{R},g}^{(r-1)} {\tilde{\nabla}_{\mathrm{R},g}^{(r-1)\mathsf{H}}}\bigr)}. % $\mathcal{O}(2 N_\mathrm{S}^3 + N_\mathrm{S}^2 + 2 N_\mathrm{S})$
			\label{eq:parameter_cg_geodesic}
		\end{equation}
		\begin{equation}
			{\mathbf{D}}_g^{(r)} = \tilde{\nabla}_{\mathrm{R},g}^{(r)} + \tilde{\gamma}_g^{(r)} {\mathbf{D}}_g^{(r-1)},
			\label{eq:direction_cg_geodesic}
		\end{equation}
		The solution can thus be updated along the geodesic in a multiplicative rotational manner
		\begin{equation}
			\mathbf{\Theta}_g^{(r+1)} = \mathbf{G}_g^{(r)}(\mu) = \exp(\mu \mathbf{D}_g^{(r)}) \mathbf{\Theta}_g^{(r)},
			\label{eq:update_geodesic}
		\end{equation}
		where an appropriate $\mu$ may be obtained by the Armijo rule.
		To double the step size, one can simply square the rotation matrix instead of recomputing the matrix exponential, since $\exp^2(\mu \mathbf{D}_g^{(r)}) = \exp(2 \mu \mathbf{D}_g^{(r)})$.

		\setalgorithmcaptionfont{\small}
		\begin{algorithm}[!t]
			\small
			\caption{Group-wise geodesic \gls{rcg} for \gls{bd}-\gls{ris} design}
			\label{ag:rcg}
			\begin{algorithmic}[1]
				\Require $f(\mathbf{\Theta})$, $G$
				\Ensure $\mathbf{\Theta}^\star$
				\Initialize {$r \gets 0$, $\mathbf{\Theta}^{(0)}$}
				\Repeat
					\For {$g \gets 1$ to $G$}
						\State $\nabla_{\mathrm{E},g}^{(r)} \gets$ \eqref{eq:gradient_euclidean} \label{ln:gradient_euclidean}
						\State $\tilde{\nabla}_{\mathrm{R},g}^{(r)} \gets$ \eqref{eq:gradient_translated}
						\State $\tilde{\gamma}_g^{(r)} \gets$ \eqref{eq:parameter_cg_geodesic}
						\State $\mathbf{D}_g^{(r)} \gets$ \eqref{eq:direction_cg_geodesic}
						\If {$\Re\bigl\{\mathrm{tr}({\mathbf{D}_g^{(r)\mathsf{H}}} \tilde{\nabla}_{\mathrm{R},g}^{(r)})\bigr\} < 0$} \Comment{not an ascent direction}
							\State $\mathbf{D}_g^{(r)} \gets \nabla_{\mathrm{R},g}^{(r)}$
						\EndIf
						\State $\mu \gets 1$
						\State $\mathbf{G}_g^{(r)}(\mu) \gets$ \eqref{eq:geodesic_translated}
						\While {$f\bigl(\mathbf{G}_g^{(r)}(2\mu)\bigr) - f(\mathbf{\Theta}_g^{(r)}) \ge \mu \cdot \mathrm{tr}(\mathbf{D}_g^{(r)} {\mathbf{D}_g^{(r)\mathsf{H}}}) / 2$}
							\State $\mu \gets 2 \mu$
						\EndWhile
						\While {$f\bigl(\mathbf{G}_g^{(r)}(\mu)\bigr) - f(\mathbf{\Theta}_g^{(r)}) < \mu / 2 \cdot \mathrm{tr}(\mathbf{D}_g^{(r)} {\mathbf{D}_g^{(r)\mathsf{H}}}) / 2$}
							\State $\mu \gets \mu / 2$
						\EndWhile
						\State $\mathbf{\Theta}_g^{(r+1)} \gets$ \eqref{eq:update_geodesic}
					\EndFor
					\State $r \gets r+1$
				\Until $\lvert f(\mathbf{\Theta}^{(r)}) - f(\mathbf{\Theta}^{(r-1)}) \rvert / f(\mathbf{\Theta}^{(r-1)}) \le \epsilon$
			\end{algorithmic}
		\end{algorithm}

		Algorithm~\ref{ag:rcg} summarizes the proposed \gls{bd}-\gls{ris} design framework based on group-wise geodesic \gls{rcg}.
		Compared to the non-geodesic approach, our method leverages the properties of unitary group to replace the add-then-retract update \eqref{eq:add_then_retract} with a multiplicative rotational update \eqref{eq:update_geodesic} along the geodesic.
		This leads to faster convergence and simplifies the step size tuning thanks to an appropriate parameter space.

		Both approaches can be performed in parallel for all groups to facilitate large-scale \gls{bd}-\gls{ris} optimization problems.
		Since block-unitary matrices are also closed under multiplication, one can avoid group-wise updates by directly operating on $\mathbf{\Theta}$ and pinching (i.e., keeping the block diagonal and nulling the other entries) the Euclidean gradient \eqref{eq:gradient_euclidean}, at the cost of higher computational complexity and slower convergence.
	\end{subsection}
\end{section}

@@

	\begin{subsection}{Pareto Frontier Characterization}\label{sc:pareto_frontier}
		We then characterize the Pareto frontier of singular values of a general $N_\mathrm{T} \times N_\mathrm{S} \times N_\mathrm{R}$ channel \eqref{eq:channel_equivalent} by maximizing their weighted sum
		\begin{maxi!}
			{\scriptstyle{\mathbf{\Theta}}}{\sum_n \rho_n \sigma_n(\mathbf{H})}{\label{op:pareto}}{\label{ob:pareto}}
			\addConstraint{\mathbf{\Theta}_g^\mathsf{H} \mathbf{\Theta}_g=\mathbf{I},}{\quad \forall g,}{\label{co:pareto_unitary}}
		\end{maxi!}
		where $n \in \mathcal{N} \triangleq \{1,\ldots,N\}$, $N \triangleq \min(N_\mathrm{T}, N_\mathrm{R})$ is the maximum channel rank, and $\rho_n$ is the weight of the $n$-th singular value that can be positive, zero, or negative.
		Varying $\{\rho_n\}_{n \in \mathcal{N}}$ characterizes the Pareto frontier that encloses the entire achievable singular value region.
		Thus, we claim the singular value Pareto problem \eqref{op:pareto} generalizes most relevant channel shaping problems.
		It can be solved by the proposed group-wise geodesic \gls{rcg} method with the following gradient expression.

		\begin{lemma}\label{lm:pareto_gradient}
			The Euclidean gradient of \eqref{ob:pareto} with respect to \gls{bd}-\gls{ris} group $g$ is
			\begin{equation}
				\frac{\partial \sum_n \rho_n \sigma_n(\mathbf{H})}{\partial \mathbf{\Theta}_g^*} = \mathbf{H}_{\mathrm{B},g}^\mathsf{H} \mathbf{U} \mathrm{diag}(\rho_1,\ldots,\rho_N) \mathbf{V}^\mathsf{H} \mathbf{H}_{\mathrm{F},g}^\mathsf{H},
				\label{eq:pareto_gradient}
			\end{equation}
			where $\mathbf{U}$ and $\mathbf{V}$ are the left and right singular matrices of $\mathbf{H}$, respectively.
		\end{lemma}
		\begin{proof}
			Please refer to Appendix~\ref{ap:pareto_gradient}.
		\end{proof}

		Algorithm~\ref{ag:rcg} can thus be invoked for problem \eqref{op:pareto} where line \ref{ln:gradient_euclidean} uses \eqref{eq:pareto_gradient} explicitly.
		With proper initialization, it is guaranteed to converge to (at least local) optimal points.
	\end{subsection}

@@

\begin{section}{Power Gain and Achievable Rate Maximization}\label{sc:power_rate}
	\begin{subsection}{Channel Power Gain}\label{sc:power}
		The \gls{mimo} channel power gain maximization problem is formulated with respect to the \gls{bd}-\gls{ris} scattering matrix
		\begin{maxi!}
			{\scriptstyle{\mathbf{\Theta}}}{\lVert \mathbf{H}_\mathrm{D} + \mathbf{H}_\mathrm{B} \mathbf{\Theta} \mathbf{H}_\mathrm{F} \rVert _\mathrm{F}^2}{\label{op:power_passive}}{\label{ob:power_passive}}
			\addConstraint{\mathbf{\Theta}_g^\mathsf{H} \mathbf{\Theta}_g=\mathbf{I}, \quad \forall g,}{}{}
		\end{maxi!}
		which generalizes the case of \gls{siso} \cite{Shen2020a}, \gls{miso} \cite{Santamaria2023,Fang2023}, single-stream \gls{mimo} \cite{Nerini2023,Nerini2023b}, and direct link-blocked \gls{mimo} with unitary \gls{ris} \eqref{eq:scattering_fc_optimal}.
		The key of solving \eqref{op:power_passive} is to balance the additive and multiplicative subspace alignments.
		Interestingly, in terms of maximizing the inner product $\langle \mathbf{H}_\mathrm{D}, \mathbf{H}_\mathrm{B} \mathbf{\Theta} \mathbf{H}_\mathrm{F} \rangle$, \eqref{op:power_passive} is reminiscent of the weighted orthogonal Procrustes problem \cite{Gower2004}
		\begin{mini!}
			{\scriptstyle{\mathbf{\Theta}}}{\lVert \mathbf{H}_\mathrm{D} - \mathbf{H}_\mathrm{B} \mathbf{\Theta} \mathbf{H}_\mathrm{F} \rVert _\mathrm{F}^2}{\label{op:weighted_orthogonal_procrustes}}{}
			\addConstraint{\mathbf{\Theta}^\mathsf{H} \mathbf{\Theta}=\mathbf{I},}{\label{cs:power_block_unitary}}{}
		\end{mini!}
		which relaxes the block-unitary constraint \eqref{cs:power_block_unitary} to unitary but still has no trivial solution.
		One lossy transformation exploits the Moore-Penrose inverse and moves $\mathbf{\Theta}$ to one side of the product \cite{Bell2003}, formulating two standard orthogonal Procrustes problems
		\begin{mini!}
			{\scriptstyle{\mathbf{\Theta}}}{\lVert \mathbf{H}_\mathrm{B}^\dagger \mathbf{H}_\mathrm{D} - \mathbf{\Theta} \mathbf{H}_\mathrm{F} \rVert _\mathrm{F}^2 \text{ or } \lVert \mathbf{H}_\mathrm{D} \mathbf{H}_\mathrm{F}^\dagger - \mathbf{H}_\mathrm{B} \mathbf{\Theta} \rVert _\mathrm{F}^2}{\label{op:standard_orthogonal_procrustes}}{}
			\addConstraint{\mathbf{\Theta}^\mathsf{H} \mathbf{\Theta}=\mathbf{I},}{}{}
		\end{mini!}
		which have global optimal solutions
		\begin{equation}
			\mathbf{\Theta} = \mathbf{U} \mathbf{V}^\mathsf{H},
			\label{eq:orthogonal_procrustes_solution}
		\end{equation}
		where $\mathbf{U}$ and $\mathbf{V}$ are respectively the left and right singular matrices of $\mathbf{H}_\mathrm{B}^\dagger \mathbf{H}_\mathrm{D} \mathbf{H}_\mathrm{F}^\mathsf{H}$ or $\mathbf{H}_\mathrm{B}^\mathsf{H} \mathbf{H}_\mathrm{D} \mathbf{H}_\mathrm{F}^\dagger$ \cite{Golub2013}.
		We emphasize that \eqref{eq:scattering_fc_optimal} and \eqref{eq:orthogonal_procrustes_solution} are valid unitary \gls{ris} solutions to \eqref{op:power_passive} when the direct link is absent and present, but the latter is neither optimal nor a generalization of the former due to the lossy transformation.

@@

		\begin{proposition}\label{pp:power}
			Starting from any feasible $\mathbf{\Theta}^{(0)}$, the sequence
			\begin{equation}
				\mathbf{\Theta}_g^{(r+1)} = \mathbf{U}_g^{(r)} \mathbf{V}_g^{(r)}, \quad \forall g.
				\label{eq:scattering_power}
			\end{equation}
			converges to a stationary point of \eqref{op:power_passive}, where $\mathbf{U}_g^{(r)}$ and $\mathbf{V}_g^{(r)}$ are the left and right compact singular matrix of
			\begin{equation}
				\mathbf{M}_g^{(r)} = \mathbf{H}_{\mathrm{B},g}^\mathsf{H} \Bigl(\mathbf{H}_\mathrm{D} + \mathbf{H}_\mathrm{B} \mathrm{diag}\bigl(\mathbf{\Theta}_{[1:g-1]}^{(r+1)},\mathbf{\Theta}_{[g:G]}^{(r)}\bigr) \mathbf{H}_\mathrm{F}\Bigr) \mathbf{H}_{\mathrm{F},g}^\mathsf{H}
				\label{eq:auxiliary_matrix_power}
			\end{equation}
		\end{proposition}

		\begin{proof}
			Please refer to Appendix~\ref{ap:power}.
		\end{proof}
	\end{subsection}

>>>>

	% 	\item What is the entire singular value region of a \gls{mimo} channel with \gls{ris}?

@@

	% pioneer
	Third, we quantify the capability of a \gls{bd}-\gls{ris} to redistribute the singular values of a point-to-point \gls{mimo} channel.
	The {Pareto frontiers} are characterized by optimizing the {weighted sum of singular values}, where the weights can be positive, zero, or negative.
	This problem is solved by the proposed geodesic \gls{rcg} algorithm.
	The resulting singular value region generalizes most relevant metrics and provides an intuitive channel shaping benchmark.
	% We also and derive some analytical singular value bounds for rank-deficient \gls{mimo} and unitary \gls{ris}.
	To explore the shaping limits of different \gls{ris} architectures, we also derive individual and collective singular value bounds for rank-deficient channels and unitary \gls{ris} via matrix analysis.
	Results validate those bounds and show that increasing \gls{bd}-\gls{ris} group size significantly enlarges the singular value region, providing wider dynamic range and better trade-off.
	% dynamic range and trade-off of singular values are significantly improved by \gls{bd}-\gls{ris}, especially in large-scale \gls{mimo} systems.
	This is the first work to comprehensively answer the channel shaping capability question from both numerical and analytical perspectives in the most general \gls{bd}-\gls{ris} setup.
	% from a Pareto perspective.

@@

		A Lie group is simultaneously a continuous group and a differentiable manifold.
		Lie algebra refers to the tangent space of the Lie group at the identity element.
		The exponential map acts as a bridge between the Lie algebra and Lie group, which allows one to recapture the local group structure using linear algebra techniques.
		The set of unitary matrices $\mathbb{U}^{L \times L}$ forms a Lie group $U(L)$ under multiplication, and the corresponding Lie algebra $\mathfrak{u}(L) \triangleq \mathcal{T}_{\mathbf{I}}\mathbb{U}^{L \times L} = \{\mathbf{M} \in \mathbb{C}^{L \times L} \mid \mathbf{M}^\mathsf{H} + \mathbf{M} = \mathbf{0}\}$ consists of skew-Hermitian matrices.
		A geodesic emanating from the identity with velocity $\mathbf{D} \in \mathfrak{u}(L)$ can be described by \cite{Edelman1998}
		\begin{equation}
			\mathbf{G}_\mathbf{I}(\mu) = \exp(\mu \mathbf{D}),
			\label{eq:geodesic_identity}
		\end{equation}
		where $\exp(\mathbf{A}) = \sum_{k=0}^\infty (\mathbf{A}^k/k!)$ is the matrix exponential and $\mu$ is the step size (i.e., magnitude of the tangent vector).
		Note that the right translation is an isometry in $U(L)$.
		During the optimization of group $g$, the geodesic evaluated at the identity \eqref{eq:geodesic_identity} should be translated to $\mathbf{\Theta}_g^{(r)}$ for successive updates \cite{Abrudan2008}
		\begin{equation}
			\mathbf{G}_g^{(r)}(\mu) = \mathbf{G}_\mathbf{I}(\mu) \mathbf{\Theta}_g^{(r)} = \exp(\mu \mathbf{D}_g^{(r)}) \mathbf{\Theta}_g^{(r)},
			\label{eq:geodesic_translated}
		\end{equation}
		while the Riemannian gradient evaluated at $\mathbf{\Theta}_g^{(r)}$ \eqref{eq:gradient_riemannian} should be translated back to the identity for exploiting the Lie algebra \cite{Abrudan2008}
		\begin{equation}
			\tilde{\nabla}_{\mathrm{R},g}^{(r)} = \nabla_{\mathrm{R},g}^{(r)} \mathbf{\Theta}_g^{(r)\mathsf{H}} = \nabla_{\mathrm{E},g}^{(r)} \mathbf{\Theta}_g^{(r)\mathsf{H}} - \mathbf{\Theta}_g^{(r)} {\nabla_{\mathrm{E},g}^{(r)\mathsf{H}}}.
			% $\mathcal{O}(2 L^3)$
			\label{eq:gradient_translated}
		\end{equation}
		After gradient translation, the deviation parameter and conjugate direction can be determined similarly to \eqref{eq:parameter_cg} and \eqref{eq:direction_cg}
		\begin{equation}
			\tilde{\gamma}_g^{(r)} = \frac{\mathrm{tr}\bigl((\tilde{\nabla}_{\mathrm{R},g}^{(r)} - \tilde{\nabla}_{\mathrm{R},g}^{(r-1)}) {\tilde{\nabla}_{\mathrm{R},g}^{(r)\mathsf{H}}}\bigr)}{\mathrm{tr}\bigl(\tilde{\nabla}_{\mathrm{R},g}^{(r-1)} {\tilde{\nabla}_{\mathrm{R},g}^{(r-1)\mathsf{H}}}\bigr)}. % $\mathcal{O}(2 N_\mathrm{S}^3 + N_\mathrm{S}^2 + 2 N_\mathrm{S})$
			\label{eq:parameter_cg_geodesic}
		\end{equation}
		\begin{equation}
			{\mathbf{D}}_g^{(r)} = \tilde{\nabla}_{\mathrm{R},g}^{(r)} + \tilde{\gamma}_g^{(r)} {\mathbf{D}}_g^{(r-1)},
			\label{eq:direction_cg_geodesic}
		\end{equation}
		The solution can thus be updated along the geodesic in a multiplicative rotational manner
		\begin{equation}
			\mathbf{\Theta}_g^{(r+1)} = \mathbf{G}_g^{(r)}(\mu) = \exp(\mu \mathbf{D}_g^{(r)}) \mathbf{\Theta}_g^{(r)},
			\label{eq:update_geodesic}
		\end{equation}
		where an appropriate $\mu$ may be obtained by the Armijo rule.
		To double the step size, one can simply square the rotation matrix instead of recomputing the matrix exponential, since $\exp^2(\mu \mathbf{D}_g^{(r)}) = \exp(2 \mu \mathbf{D}_g^{(r)})$.

		\setalgorithmcaptionfont{\small}
		\begin{algorithm}[!t]
			\small
			\caption{Group-wise geodesic \gls{rcg} for \gls{bd}-\gls{ris} design}
			\label{ag:rcg}
			\begin{algorithmic}[1]
				\Require $f(\mathbf{\Theta})$, $G$
				\Ensure $\mathbf{\Theta}^\star$
				\Initialize {$r \gets 0$, $\mathbf{\Theta}^{(0)}$}
				\Repeat
					\For {$g \gets 1$ to $G$}
						\State $\nabla_{\mathrm{E},g}^{(r)} \gets$ \eqref{eq:gradient_euclidean} \label{ln:gradient_euclidean}
						\State $\tilde{\nabla}_{\mathrm{R},g}^{(r)} \gets$ \eqref{eq:gradient_translated}
						\State $\tilde{\gamma}_g^{(r)} \gets$ \eqref{eq:parameter_cg_geodesic}
						\State $\mathbf{D}_g^{(r)} \gets$ \eqref{eq:direction_cg_geodesic}
						\If {$\Re\bigl\{\mathrm{tr}({\mathbf{D}_g^{(r)\mathsf{H}}} \tilde{\nabla}_{\mathrm{R},g}^{(r)})\bigr\} < 0$} \Comment{not an ascent direction}
							\State $\mathbf{D}_g^{(r)} \gets \tilde{\nabla}_{\mathrm{R},g}^{(r)}$
						\EndIf
						\State $\mu \gets 1$
						\State $\mathbf{G}_g^{(r)}(\mu) \gets$ \eqref{eq:geodesic_translated}
						\While {$f\bigl(\mathbf{G}_g^{(r)}(2\mu)\bigr) - f(\mathbf{\Theta}_g^{(r)}) \ge \mu \cdot \mathrm{tr}(\mathbf{D}_g^{(r)} {\mathbf{D}_g^{(r)\mathsf{H}}}) / 2$} \label{ln:armijo_start}
							\State $\mu \gets 2 \mu$
						\EndWhile
						\While {$f\bigl(\mathbf{G}_g^{(r)}(\mu)\bigr) - f(\mathbf{\Theta}_g^{(r)}) < \mu / 2 \cdot \mathrm{tr}(\mathbf{D}_g^{(r)} {\mathbf{D}_g^{(r)\mathsf{H}}}) / 2$}
							\State $\mu \gets \mu / 2$
						\EndWhile \label{ln:armijo_end}
						\State $\mathbf{\Theta}_g^{(r+1)} \gets$ \eqref{eq:update_geodesic}
					\EndFor
					\State $r \gets r+1$
				\Until $\lvert f(\mathbf{\Theta}^{(r)}) - f(\mathbf{\Theta}^{(r-1)}) \rvert / f(\mathbf{\Theta}^{(r-1)}) \le \epsilon$
			\end{algorithmic}
		\end{algorithm}

		Algorithm~\ref{ag:rcg} summarizes the proposed \gls{bd}-\gls{ris} design framework based on group-wise geodesic \gls{rcg}.
		Compared to the non-geodesic approach, it leverages the Lie group properties to replace the add-then-retract update \eqref{eq:add_then_retract} with a multiplicative rotational update \eqref{eq:update_geodesic} along the geodesic.
		This leads to faster convergence and simplifies the step size tuning thanks to appropriate parameter space.
		% Despite the difference in
		Convergence to a local optimum is still guaranteed if not initialized at a stationary point.
		Note that the group-wise updates can be performed in parallel to facilitate large-scale \gls{bd}-\gls{ris} optimization problems.
		Since block-unitary matrices are also closed under multiplication, one can avoid group-wise updates by directly operating on $\mathbf{\Theta}$ and pinching (i.e., keeping the block diagonal and nulling the other entries) the Euclidean gradient \eqref{eq:gradient_euclidean}, with potentially higher computational complexity and slower convergence.
	\end{subsection}
\end{section}

@@

	\begin{subsection}{Pareto Frontier Characterization}\label{sc:pareto_frontier}
		We then characterize the Pareto frontier of singular values of a general $N_\mathrm{T} \times N_\mathrm{S} \times N_\mathrm{R}$ channel \eqref{eq:channel_equivalent} by maximizing their weighted sum
		\begin{maxi!}
			{\scriptstyle{\mathbf{\Theta}}}{\sum_n \rho_n \sigma_n(\mathbf{H})}{\label{op:pareto}}{\label{ob:pareto}}
			\addConstraint{\mathbf{\Theta}_g^\mathsf{H} \mathbf{\Theta}_g=\mathbf{I},}{\quad \forall g,}{\label{co:pareto_unitary}}
		\end{maxi!}
		where $n \in \mathcal{N} \triangleq \{1,\ldots,N\}$, $N \triangleq \min(N_\mathrm{T}, N_\mathrm{R})$ is the maximum channel rank, and $\rho_n$ is the weight of the $n$-th singular value that can be positive, zero, or negative.
		Varying $\{\rho_n\}_{n \in \mathcal{N}}$ characterizes the Pareto frontier that encloses the entire singular value region.
		Thus, we claim problem \eqref{op:pareto} generalizes most singular value shaping problems.
		It can be solved optimally by Algorithm~\ref{ag:rcg} with the Euclidean gradient given by Lemma \ref{lm:pareto_gradient}.
		% It can be solved optimally by Algorithm~\ref{ag:rcg} where line \ref{ln:gradient_euclidean} uses the following Euclidean gradient expression \eqref{eq:pareto_gradient}.

		\begin{lemma}\label{lm:pareto_gradient}
			The Euclidean gradient of \eqref{ob:pareto} with respect to \gls{bd}-\gls{ris} group $g$ is
			\begin{equation}
				\frac{\partial \sum_n \rho_n \sigma_n(\mathbf{H})}{\partial \mathbf{\Theta}_g^*} = \mathbf{H}_{\mathrm{B},g}^\mathsf{H} \mathbf{U} \mathrm{diag}(\rho_1,\ldots,\rho_N) \mathbf{V}^\mathsf{H} \mathbf{H}_{\mathrm{F},g}^\mathsf{H},
				\label{eq:pareto_gradient}
			\end{equation}
			where $\mathbf{U}$ and $\mathbf{V}$ are the left and right compact singular matrices of $\mathbf{H}$, respectively.
		\end{lemma}
		\begin{proof}
			Please refer to Appendix~\ref{ap:pareto_gradient}.
		\end{proof}

		We then analyze the computational complexity of solving Pareto singular value problem \eqref{op:pareto} by Algorithm~\ref{ag:rcg}.
		To update each \gls{bd}-\gls{ris} group, compact \gls{svd} of $\mathbf{H}$ requires $\mathcal{O}(N N_\mathrm{T} N_\mathrm{R})$, Euclidean gradient \eqref{eq:pareto_gradient} requires $\mathcal{O}\bigl(L N (N_\mathrm{T}+N_\mathrm{R}+L+1) \bigr)$, Riemannian gradient translation \eqref{eq:gradient_translated} requires $\mathcal{O}(L^3)$, deviation parameter \eqref{eq:parameter_cg_geodesic} and conjugate direction \eqref{eq:direction_cg_geodesic} together require $\mathcal{O}(L^2)$, and matrix exponential \eqref{eq:update_geodesic} requires $\mathcal{O}(L^3)$ operations \cite{Moler2003}.
		The overall complexity is thus $\mathcal{O}\bigl(I_\text{RCG} G \bigl(N N_\mathrm{T} N_\mathrm{R} + L N (N_\mathrm{T}+N_\mathrm{R}+L+1) + I_\text{BLS} L^3\bigr)\bigr)$, where $I_\text{RCG}$ and $I_\text{BLS}$ are the number of iterations for geodesic \gls{rcg} and backtracking line search (line \ref{ln:armijo_start}--\ref{ln:armijo_end} of Algorithm \ref{ag:rcg}), respectively.
	\end{subsection}

@@

\begin{section}{Power Gain and Achievable Rate Maximization}\label{sc:power_rate}
	\begin{subsection}{Channel Power Gain}\label{sc:power}
		The \gls{mimo} channel power gain maximization problem is formulated with respect to the \gls{bd}-\gls{ris} scattering matrix
		\begin{maxi!}
			{\scriptstyle{\mathbf{\Theta}}}{\lVert \mathbf{H}_\mathrm{D} + \mathbf{H}_\mathrm{B} \mathbf{\Theta} \mathbf{H}_\mathrm{F} \rVert _\mathrm{F}^2}{\label{op:power_passive}}{\label{ob:power_passive}}
			\addConstraint{\mathbf{\Theta}_g^\mathsf{H} \mathbf{\Theta}_g=\mathbf{I}, \quad \forall g,}{}{}
		\end{maxi!}
		which generalizes the case of \gls{siso} \cite{Shen2020a}, \gls{miso} \cite{Santamaria2023,Fang2023}, single-stream \gls{mimo} \cite{Nerini2023,Nerini2023b}, and direct link-blocked \gls{mimo} with unitary \gls{ris} \eqref{eq:scattering_fc_optimal}.
		The key of solving \eqref{op:power_passive} is to balance the additive and multiplicative subspace alignments.
		Interestingly, in terms of maximizing the inner product $\langle \mathbf{H}_\mathrm{D}, \mathbf{H}_\mathrm{B} \mathbf{\Theta} \mathbf{H}_\mathrm{F} \rangle$, \eqref{op:power_passive} is reminiscent of the weighted orthogonal Procrustes problem \cite{Gower2004}
		\begin{mini!}
			{\scriptstyle{\mathbf{\Theta}}}{\lVert \mathbf{H}_\mathrm{D} - \mathbf{H}_\mathrm{B} \mathbf{\Theta} \mathbf{H}_\mathrm{F} \rVert _\mathrm{F}^2}{\label{op:weighted_orthogonal_procrustes}}{}
			\addConstraint{\mathbf{\Theta}^\mathsf{H} \mathbf{\Theta}=\mathbf{I},}{\label{cs:power_block_unitary}}{}
		\end{mini!}
		which relaxes the block-unitary constraint \eqref{cs:power_block_unitary} to unitary but still has no trivial solution.
		One lossy transformation exploits the Moore-Penrose inverse and moves $\mathbf{\Theta}$ to one side of the product \cite{Bell2003}, formulating two standard orthogonal Procrustes problems
		\begin{mini!}
			{\scriptstyle{\mathbf{\Theta}}}{\lVert \mathbf{H}_\mathrm{B}^\dagger \mathbf{H}_\mathrm{D} - \mathbf{\Theta} \mathbf{H}_\mathrm{F} \rVert _\mathrm{F}^2 \text{ or } \lVert \mathbf{H}_\mathrm{D} \mathbf{H}_\mathrm{F}^\dagger - \mathbf{H}_\mathrm{B} \mathbf{\Theta} \rVert _\mathrm{F}^2}{\label{op:standard_orthogonal_procrustes}}{}
			\addConstraint{\mathbf{\Theta}^\mathsf{H} \mathbf{\Theta}=\mathbf{I},}{}{}
		\end{mini!}
		which have global optimal solutions
		\begin{equation}
			\mathbf{\Theta} = \mathbf{U} \mathbf{V}^\mathsf{H},
			\label{eq:orthogonal_procrustes_solution}
		\end{equation}
		where $\mathbf{U}$ and $\mathbf{V}$ are respectively the left and right compact singular matrices of $\mathbf{H}_\mathrm{B}^\dagger \mathbf{H}_\mathrm{D} \mathbf{H}_\mathrm{F}^\mathsf{H}$ or $\mathbf{H}_\mathrm{B}^\mathsf{H} \mathbf{H}_\mathrm{D} \mathbf{H}_\mathrm{F}^\dagger$ \cite{Golub2013}.
		We emphasize that \eqref{eq:scattering_fc_optimal} and \eqref{eq:orthogonal_procrustes_solution} are valid unitary \gls{ris} solutions to \eqref{op:power_passive} when the direct link is absent and present, but the latter is neither optimal nor a generalization of the former due to the lossy transformation.

@@

		\begin{proposition}\label{pp:power}
			Starting from any feasible $\mathbf{\Theta}^{(0)}$, the sequence
			\begin{equation}
				\mathbf{\Theta}_g^{(r+1)} = \mathbf{U}_g^{(r)} \mathbf{V}_g^{(r)}, \quad \forall g.
				\label{eq:scattering_power}
			\end{equation}
			converges to a stationary point of \eqref{op:power_passive}, where $\mathbf{U}_g^{(r)}$ and $\mathbf{V}_g^{(r)}$ are the left and right compact singular matrices of
			\begin{equation}
				\mathbf{M}_g^{(r)} = \mathbf{H}_{\mathrm{B},g}^\mathsf{H} \Bigl(\mathbf{H}_\mathrm{D} + \mathbf{H}_\mathrm{B} \mathrm{diag}\bigl(\mathbf{\Theta}_{[1:g-1]}^{(r+1)},\mathbf{\Theta}_{[g:G]}^{(r)}\bigr) \mathbf{H}_\mathrm{F}\Bigr) \mathbf{H}_{\mathrm{F},g}^\mathsf{H}
				\label{eq:auxiliary_matrix_power}
			\end{equation}
		\end{proposition}

		\begin{proof}
			Please refer to Appendix~\ref{ap:power}.
		\end{proof}
